---
title: "Group Assignment 1"
output: pdf_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
### Setup Libraries
library(stats)
library(dplyr)
library(cluster)
library(ggplot2)
library(scales)
library(cluster)
library(factoextra)
options(warn = -1)
```

# Reading Data
```{r}
# import dataset
customer = read.csv("Wholesale customers data.csv")

# view data
head(customer)

# summary of data for the six categories
summary(customer[,2:8])
```
# Data Exploration and Processing 


## Normalization
### We want to normalzie the data to first check for outliers as it will help in
### getting better visuals in box plots.
```{r}
# declare normalize function using min-max method
normalize = function(x) {
  return((x - min(x)) / (max(x) - min(x)))
}

# normalize iris dataset from columns 1 to 4
customer_normalized = customer %>% mutate_at(c(3:8), normalize)
summary(customer_normalized[,3:8])
```


## Boxplots
```{r}
# boxplots for each feature
ggplot(data = customer_normalized, aes(x = "", y = Fresh)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 1)) +
  ggtitle("Boxplot of Fresh")

ggplot(data = customer_normalized, aes(x = "", y = Milk)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 1)) +
  ggtitle("Boxplot of Milk")

ggplot(data = customer_normalized, aes(x = "", y = Grocery)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 1)) +
  ggtitle("Boxplot of Grocery")

ggplot(data = customer_normalized, aes(x = "", y = Frozen)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 1)) +
  ggtitle("Boxplot of Frozen")

ggplot(data = customer_normalized, aes(x = "", y = Detergents_Paper)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 1)) +
  ggtitle("Boxplot of Detergents_Paper")

ggplot(data = customer_normalized, aes(x = "", y = Delicatessen)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 1)) +
  ggtitle("Boxplot of Delicatessen")
```

### We can certainly see that there are outliers in the data(in each feature  
### column) that need to treated/removed.


## Outlier Flagging
### We want to flag outliers on a row level, i.e., for each client and check the 
### distribution of outliers for each feature colum among all the clients.
```{r}
# create list of outlier data points for each feature column
fresh_outliers <- boxplot(customer$Fresh, plot=FALSE)$out
milk_outliers <- boxplot(customer$Milk, plot=FALSE)$out
grocery_outliers <- boxplot(customer$Grocery, plot=FALSE)$out
frozen_outliers <- boxplot(customer$Frozen, plot=FALSE)$out
detergent_outliers <- boxplot(customer$Detergents_Paper, plot=FALSE)$out
delicate_outliers <- boxplot(customer$Delicatessen, plot=FALSE)$out

# create an outlier flag based on the lists created above
customer$Fresh_Outlier_Flag <- ifelse(customer$Fresh %in% fresh_outliers, 1, 0)
customer$Milk_Outlier_Flag <- ifelse(customer$Milk %in% milk_outliers, 1, 0)
customer$Grocery_Outlier_Flag <- ifelse(customer$Grocery %in% grocery_outliers
                                        , 1, 0)
customer$Frozen_Outlier_Flag <- ifelse(customer$Frozen %in% frozen_outliers
                                       , 1, 0)
customer$Detergent_Outlier_Flag <- ifelse(customer$Detergents_Paper %in% 
                                            detergent_outliers, 1, 0)
customer$Delicate_Outlier_Flag <- ifelse(customer$Delicatessen %in% 
                                           delicate_outliers, 1, 0)

# create a total outlier column containing total outliers per client across
# all feature columns
customer$Total_Outliers <- customer$Fresh_Outlier_Flag + 
  customer$Milk_Outlier_Flag + customer$Grocery_Outlier_Flag + 
  customer$Frozen_Outlier_Flag + customer$Detergent_Outlier_Flag + 
  customer$Delicate_Outlier_Flag

head(customer[,9:15])
```

## Outlier Summary
```{r}
# get outlier percentages for each feature column
cat("Total Rows: ",nrow(customer), "\n")
cat("Total Outliers: ",nrow(customer %>% filter(Total_Outliers>=1)), "\n\n")

cat("Outliers Percentage\n")
cat("-------------------\n")

# Total
cat("Total: ", percent(nrow(customer %>% filter(Total_Outliers>=1))
                       / nrow(customer)), "\n")

# Fresh
cat("Fresh: ", percent(nrow(customer %>% filter(Fresh_Outlier_Flag==1))
                       / nrow(customer)), "\n")

# Milk
cat("Milk: ", percent(nrow(customer %>% filter(Milk_Outlier_Flag==1))
                      / nrow(customer)), "\n")

# Grocery
cat("Grocery: ", percent(nrow(customer %>% filter(Grocery_Outlier_Flag==1))
                         / nrow(customer)), "\n")

# Frozen
cat("Frozen: ", percent(nrow(customer %>%
                               filter(Frozen_Outlier_Flag==1))
                        / nrow(customer)), "\n")

# Detergent
cat("Detergent Paper: ", percent(nrow(customer %>%
                                        filter(Detergent_Outlier_Flag==1))
                                 / nrow(customer)), "\n")

# Delicatessen
cat("Delicatessen: ", percent(nrow(customer %>%
                                     filter(Delicate_Outlier_Flag==1))
                              / nrow(customer)), "\n")
```

```{r}
# get how many clients have 0,1,2.. outliers
customer %>% group_by(Total_Outliers) %>% summarise(Count = n())
```

### We can see that out of 440 clients, 332 don't have the outliers present in 
### the dataset. Our clustering approach will start with taking data with no
### outliers and then we will start including clients with 1/2/3.. outlier per  
### row to see if results drastically change. We will perform a different  
### exploration exercise on the remaining outlier data to check for   
### meaningful insights.


# Clustering 


## Remove Outliers
```{r}
# outlier dataframe
customer_out = customer %>% filter(Total_Outliers >= 2)

# removing outliers and storin in new dataframe
customer_final = customer %>% filter(Total_Outliers < 2)
head(customer_final)
```
### We tried running clustering models using k = 1,2,3,4. We observed that 
### k = 2 is gving us the best results. Hence, the above piece of code takes 
### Total Outliers < 2 to get the final dataset on which clustering will be 
### performed.


## Normalization
### Normalizing data for running clustering models using min-max method
```{r}
# normalize
customer_final_norm = customer_final %>% mutate_at(c(3:8), normalize)

# create distance matrix with euclidean distance
distance_matrix = dist(customer_final_norm[,3:8], method = "euclidean")
```


## K-Means Clustering 
### Check for k value 
```{r}
# calculating SSE
SSE_curve <- c()
for (k in 1:10) {
  kcluster = kmeans(customer_final_norm[,3:8], k)
  sse = kcluster$tot.withinss
  SSE_curve[k] = sse}

# plot SSE against number of clusters
plot(1:10, SSE_curve, type = "b", main = "SSE Curve", xlab = "k Value", ylab = "SSE")
```

### The elbow plot indicates we can take 2/3/4 clusters as the  
### sum of squared errors start to get stagnant after k = 5


### Run Model
```{r}
# run k-means clustering 
cat("For k = 2: \n")
kcluster = kmeans(customer_final_norm[,3:8], centers = 2)
sc = silhouette(kcluster$cluster, dist = distance_matrix)
summary(sc)

cat("\n For k = 3: \n")
kcluster = kmeans(customer_final_norm[,3:8], centers = 3)
sc = silhouette(kcluster$cluster, dist = distance_matrix)
summary(sc)

cat("\n For k = 4: \n")
kcluster = kmeans(customer_final_norm[,3:8], centers = 4)
sc = silhouette(kcluster$cluster, dist = distance_matrix)
summary(sc)
```
### We are observing that k=2 has the best silhouette score. Therefore, we will 
### be going ahead with k=2 for K-Means

```{r}
kcluster = kmeans(customer_final_norm[,3:8], centers = 2)

# get cluster labels into original dataframe
customer_final$cluster = kcluster$cluster

fviz_cluster(kcluster, geom = "point", data = customer_final_norm[,3:8]) + 
  ggtitle("K-Means Clustering (k=2)")
```


## Hierarchical Clustering 
### Check for k value
```{r}
# Hierarchical Clustering
hierarchical = hclust(distance_matrix, method = "ward.D")
plot(hierarchical)
rect.hclust(hierarchical, k = 2)
```

### The dendogram indicates we can take 2/3 clusters.


### Run Model
```{r}
cat("For k = 2: \n")
customer_final_norm$cluster = cutree(hierarchical, k = 2)
sc = silhouette(customer_final_norm$cluster, dist = distance_matrix)
summary(sc)

cat("For k = 3: \n")
customer_final_norm$cluster = cutree(hierarchical, k = 3)
sc = silhouette(customer_final_norm$cluster, dist = distance_matrix)
summary(sc)
```
### We are observing that k=2 has the best silhouette score. Therefore, we will 
### be going ahead with k=2 for Hierarchical

```{r}
clusters = cutree(hierarchical, k = 2)

# Visualize clusters
fviz_cluster(list(data = customer_final_norm[,3:8], geom = "point", cluster = clusters))


# Visualize the dendrogram with rectangles around clusters
fviz_dend(hierarchical, k = 2, rect = TRUE)

```


# Model Evaluation
### We compared the silhouette score from best models from K-Means(0.449) 
### and Hierarchical(0.376) clustering techniques and came to a conclusion that
### we should use K-Means with k=2 as the clusters formed are easily 
### distinguishable.


# Result Interpretation
### To intrepret the results, we are taking an average of each feature column
### across each level of data - Cluster, Region and Channel to get the patterns
### around spendings of each cluster. Using those insights, we will generate
### recommendations for the XYZ company.
```{r}
customer_cluster_summary <- 
  customer_final %>% 
  group_by(cluster,Region,Channel) %>% 
  summarise(
    Count = n(),
    Avg_Fresh = mean(Fresh),
    Avg_Milk = mean(Milk),
    Avg_Grocery = mean(Grocery),
    Avg_Frozen = mean(Frozen),
    Avg_Detergents_Paper = mean(Detergents_Paper),
    Avg_Delicatessen = mean(Delicatessen)
  )
customer_cluster_summary
#write.csv(customer_cluster_summary, "customer_cluster_summary.csv")
```

### We also want to check patterns for clients that came out as outliers during
### our analysis.
```{r}
customer_outlier_summary <- 
  customer_out %>% 
  group_by(Region,Channel) %>% 
  summarise(
    Count = n(),
    Avg_Fresh = mean(Fresh),
    Avg_Milk = mean(Milk),
    Avg_Grocery = mean(Grocery),
    Avg_Frozen = mean(Frozen),
    Avg_Detergents_Paper = mean(Detergents_Paper),
    Avg_Delicatessen = mean(Delicatessen)
  )
customer_outlier_summary
#write.csv(customer_outlier_summary, "customer_outliers_summary.csv")
```